Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-23 analysis/paper/article.qmd
minor edits, review JB's additions to the discussion section

This discrepancy in our data set—wherein advertisements for tenure-track positions dominate the wiki despite their increasing rarity in larger market—may be due to several factors. The first involves the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but are instead disseminated through local email lists and are filled by people close to the hiring department, such as recently graduated students. A second factor is bias in our data—with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track positions. Tellingly, Passalacqua’s examination of a similar wiki over a seven-year span showed a similar dominance of permanent positions—291/474 (63%) were for tenure-track hires at various levels (2018: Table 2). This congruence between subfields suggests that these patterns are a reflection of the *wiki*, rather than the market itself.

It is no secret that tenure-track jobs have traditionally been considered the "gold standard" of career outcomes for PhD students in the social sciences and humanities. As Kelsky outlines in *The Professor Is In*, her best-selling guide for PhD students pursuing faculty careers, "...most ranking graduate programs sill consider any Ph.D. who doesn't land a tenure-track job a failure or an aberration... Graduate students absorb this value system and judge themselves harshly" [-@kelsky2015, pp. 11]. These rubrics of success have been common across the humanities and social sciences for over half a century. Evaluating the demise of the National Endowment for Humanities Funded "Program to ready PhDs for Careers in Business" (CIB), which aimed to facilitate humanities PhDs' transition to the corporate world in the late 1970s, Franczak points to the dissonance between participants' incentives and outcomes:

@@

The desire to pursue an academic career also characterizes more recent PhD cohorts, including those within archaeology. Brami et al's survey found that despite the long odds and grueling precaraity, 71% of their early career respondents wished to remain in academia [=@bramiPrecariousFutureReflections2023, pp. 242]. Overall, the obsessive valorization of permanent faculty positions in the face of a precipitous decline in available jobs is indicative of the potent combination of denial and delusion inacross academia. As Cefkin and Schwegler argue "This asymmetry between the “ideal” career path and the experiences of most graduates highlights the deep and growing chasm between the future that higher education institutions envision—and strongly incentivize—and what is actually happening" [@cefkin].

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019]. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

>>>>

This discrepancy in our data set—wherein advertisements for tenure-track positions dominate the wiki despite their increasing rarity in larger market—may be due to several factors. The first involves the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but are instead disseminated through local email lists and are filled by people close to the hiring department, such as recently graduated students. A second factor is bias in our data—with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track positions. Tellingly, Passalacqua’s examination of a similar wiki over a seven-year span showed a similar dominance of permanent positions—291/474 (63%) were for tenure-track hires at various levels (2018: Table 2). This congruence between subfields suggests that these patterns are a reflection of the Wiki, rather than the market itself.

It is no secret that tenure-track jobs have traditionally been considered the "gold standard" of career outcomes for PhD students in the social sciences and humanities. As Kelsky outlines in *The Professor Is In*, her influential and best-selling guide for PhD students pursuing faculty careers, "...most ranking graduate programs sill consider any PhD who doesn't land a tenure-track job a failure or an aberration... Graduate students absorb this value system and judge themselves harshly" [-@kelsky2015, pp. 11]. These rubrics of success have been common across the humanities and social sciences for over half a century. Evaluating the demise of the National Endowment for Humanities Funded "Program to ready PhDs for Careers in Business" (CIB), which aimed to facilitate humanities PhDs' transition to the corporate world in the late 1970s, Franczak points to the dissonance between participants' incentives and outcomes:

@@

The desire to pursue an academic career also characterizes more recent PhD cohorts, including those within archaeology. Brami et al's survey found that despite the long odds and grueling precaraity, 71% of their early career respondents wished to remain in academia [-@bramiPrecariousFutureReflections2023, pp. 242]. Overall, the obsessive valorization of permanent faculty positions in the face of a precipitous decline in available jobs is indicative of the potent combination of denial and delusion inacross academia. As Cefkin and Schwegler argue "This asymmetry between the 'ideal' career path and the experiences of most graduates highlights the deep and growing chasm between the future that higher education institutions envision—and strongly incentivize—and what is actually happening" [@cefkin].

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, subjective, and relational perspectives, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019]. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.
